A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-12-11 main.tex
3.6.2, 4.1, 5 edits

We generated these synthetic phenotypes using three types of high-dimensional ($p > n$) $\bG$ matrices, which were then standardized to yield the corresponding $\bX$ matrix: (1) simulated SNP data from admixed subpopulations, (2) simulated SNP data from independent subpopulations, and (3) real genome-wide SNP data from known caucasian ($n = 417$), African American ($n = 134$), and Hispanic ($n = 37$) subpopulations \citep{larkin2015objectives}. We refer to these as the 1D Linear Admixture, Independent Subpopulations, and Empirical data, respectively. For all data types $p = 1000$ with 50 causal SNPs. For the simulated 1D Linear Admixture and Independent Subpopulations data $n = 200$ unless otherwise specified, and $n = 588$ for the Empirical data.

@@

\begin{table}[H]
\centering
\begin{tabularx}{400pt}{cclYYY}
\toprule
$\eta$ & $\xi$ & \multicolumn{1}{c}{Method} & 1D Linear Admixture & Independent Subpopulations & Empirical \\ 
\midrule
\multirow{6}{*}{0.5} & \multirow{3}{*}{0} & lasso & 0.25 (0.05) & 0.23 (0.05) & 0.37 (0.06) \\ 
& & PC-lasso & 0.23 (0.05) & 0.21 (0.05) & 0.36 (0.06) \\ 
& & LMM-lasso & 0.25 (0.05) & 0.23 (0.05) & 0.41 (0.06) \\ 
\cmidrule{2-6}
& \multirow{3}{*}{0.8} & lasso & 0.24 (0.05) & 0.26 (0.06) & 0.29 (0.06) \\ 
& & PC-lasso & 0.28 (0.05) & 0.36 (0.06) & 0.46 (0.06) \\ 
& &  LMM-lasso & 0.28 (0.05) & 0.39 (0.06) & 0.61 (0.06) \\ 
\bottomrule
\end{tabularx}
\caption{True positive rates, mean (SD), for different data types in the presence ($\xi = 0$) and absence ($\xi = 0.8$) of environmental effects. For the data sets with simulated genotypes, \pb{???} subpopulations are present.}
\label{tab:var_sel}
\end{table}

@@

LMM-lasso-ggmix does tend to provide improved estimates of $\eta$ compared to LMM-lasso-Rakitsch, particularly in the presence of lower $\eta$ values and where $n \ll p$ (\ref{fig:eta_beta_mse}, middle panel), although these improvements did not help it to estimate $\bbeta$ more accurately. In addition, at high signal-to-noise ratios ($\eta=0.8$), the performance of LMM-lasso-ggmix was somewhat erratic.

@@

Although the purpose of this article is to elucidate the manner in which failing to correct for unobserved environmental effects can bias SNP effect estimates, equation \eqref{eqn:partition_e} also highlights the manner in which such unobserved environmental effects can lead to substantial bias in narrow-sense heritability estimates using SNP data. In particular, these estimates are likely to be inflated by the quantity $\boldsymbol{tau}$ unless steps are taken to address population structure. A detailed treatment of narrow-sense heritability estimation is beyond the scope of this paper, but we hope this emphasizes the broad implications and importance of distinguishing between population structure and environmental effects.

>>>>

We generated these synthetic phenotypes using three types of high-dimensional ($p > n$) $\bG$ matrices, which were then standardized to yield the corresponding $\bX$ matrix: (1) simulated SNP data from admixed subpopulations, (2) simulated SNP data from independent subpopulations, and (3) real genome-wide SNP data from known caucasian ($n = 417$), African American ($n = 134$), and Hispanic ($n = 37$) subpopulations \citep{larkin2015objectives}. We refer to these as the 1D Linear Admixture, Independent Subpopulations, and Empirical data, respectively. For all data types $p = 1000$ with 50 causal SNPs. For the simulated 1D Linear Admixture and Independent Subpopulations data \anna{$n = 200$ unless otherwise specified}, and $n = 588$ for the Empirical data.

@@

\begin{table}[H]
\centering
\begin{tabularx}{400pt}{cclYYY}
\toprule
$\eta$ & $\xi$ & \multicolumn{1}{c}{Method} & 1D Linear Admixture & Independent Subpopulations & Empirical \\ 
\midrule
\multirow{6}{*}{0.5} & \multirow{3}{*}{0} & lasso & 0.25 (0.05) & 0.23 (0.05) & 0.37 (0.06) \\ 
& & PC-lasso & 0.23 (0.05) & 0.21 (0.05) & 0.36 (0.06) \\ 
& & LMM-lasso & 0.25 (0.05) & 0.23 (0.05) & 0.41 (0.06) \\ 
\cmidrule{2-6}
& \multirow{3}{*}{0.8} & lasso & 0.24 (0.05) & 0.26 (0.06) & 0.29 (0.06) \\ 
& & PC-lasso & 0.28 (0.05) & 0.36 (0.06) & 0.46 (0.06) \\ 
& &  LMM-lasso & 0.28 (0.05) & 0.39 (0.06) & 0.61 (0.06) \\ 
\bottomrule
\end{tabularx}
\caption{True positive rates, mean (SD), for different data types in the presence ($\xi = 0$) and absence ($\xi = 0.8$) of environmental effects. For the data sets with simulated genotypes, 4 subpopulations are present.}
\label{tab:var_sel}
\end{table}

@@

LMM-lasso-ggmix does tend to provide improved estimates of $\eta$ compared to LMM-lasso-Rakitsch, particularly in the presence of lower $\eta$ values and where $n \ll p$ (Figure \ref{fig:eta_beta_mse}, middle panel), although these improvements did not help it to estimate $\bbeta$ more accurately. In addition, at high signal-to-noise ratios ($\eta=0.8$), the performance of LMM-lasso-ggmix was somewhat erratic.

@@

Although the purpose of this article is to elucidate the manner in which failing to correct for unobserved environmental effects can bias SNP effect estimates, equation \eqref{eqn:partition_e} also highlights the manner in which such unobserved environmental effects can lead to substantial bias in narrow-sense heritability estimates using SNP data. In particular, these estimates are likely to be inflated by the quantity $\boldsymbol{\tau}$ unless steps are taken to address population structure. A detailed treatment of narrow-sense heritability estimation is beyond the scope of this paper, but we hope this emphasizes the broad implications and importance of distinguishing between population structure and environmental effects.
